First paper - Draft

### 2017-07-05 paper/paper.tex
All graphs and tables visible in tex. Rearranging and recommenting

{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}

@@

We start by look at the achievement for some countries. We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly 0.9. This pattern happens similarly for literacy with an increase of 0.6. For such a short period of time, the magnitud of the increase is reasonably big.

@@

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation. The actual difference is of about 0.6. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

Something reassuring is that mathematics and reading follow basically the same trend across all countries. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

@@

@@

@@

For all country-years pooled we see a very strong correlation but contrary to what we would expect. We find that countries with high performance have also a very low level of inequality. In fact, even if we separate it by year.

In fact, we see a strong correspondence. In countries like Germany and France we see quite a functional form between both inequality and performance. This pattern is also seen in in the UK, Austria, Australia and New Zealand. Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably. On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.

@@

In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing. We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

@@

>>>>

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Wed Jul  5 09:41:15 2017
\scalebox{0.75}{
\begin{tabular}{llllllll}
   \hline \
                          & & & Low SES & & & High SES \\
                          \cmidrule(l){3-5}
                          \cmidrule(l){6-8} Year & Countries & \% Low Edu & \% High Edu & Sample size & \% Low Edu & \% High Edu & Sample size \\ \hline2000 & Australia & 5\% & 1\% & 294 & 0\% & 97\% & 270 \\ 
  2000 & Austria & 4\% & 0\% & 294 & 0\% & 99\% & 235 \\ 
  2000 & Canada & 3\% & 1\% & 1970 & 0\% & 99\% & 1357 \\ 
  2000 & Denmark & 4\% & 0\% & 230 & 0\% & 97\% & 228 \\ 
  2000 & Finland & 2\% & 0\% & 261 & 0\% & 97\% & 267 \\ 
  2000 & France & 14\% & 0\% & 253 & 0\% & 97\% & 258 \\ 
  2000 & Germany & 2\% & 0\% & 250 & 0\% & 93\% & 303 \\ 
  2000 & Hungary & 1\% & 0\% & 291 & 0\% & 99\% & 254 \\ 
  2000 & Italy & 3\% & 0\% & 278 & 0\% & 99\% & 278 \\ 
  2000 & Netherlands & 9\% & 0\% & 129 & 0\% & 99\% & 158 \\ 
  2000 & Poland & 0\% & 0\% &  & 0\% & 92\% & 166 \\ 
  2000 & Spain & 15\% & 0\% & 318 & 0\% & 98\% & 340 \\ 
  2000 & Sweden & 4\% & 1\% & 241 & 0\% & 98\% & 246 \\ 
  2000 & United Kingdom & 2\% & 0\% & 531 & 0\% & 95\% & 415 \\ 
  2000 & United States & 5\% & 0\% & 222 & 0\% & 97\% & 164 \\ 
  2015 & Australia & 2\% & 2\% & 1694 & 0\% & 94\% & 1235 \\ 
  2015 & Austria & 4\% & 0\% & 676 & 0\% & 96\% & 705 \\ 
  2015 & Canada & 1\% & 3\% & 2215 & 0\% & 97\% & 1863 \\ 
  2015 & Denmark & 4\% & 0\% & 1013 & 0\% & 99\% & 595 \\ 
  2015 & Finland & 2\% & 1\% & 575 & 0\% & 99\% & 584 \\ 
  2015 & France & 3\% & 0\% & 570 & 0\% & 94\% & 615 \\ 
  2015 & Germany & 5\% & 0\% & 545 & 0\% & 96\% & 582 \\ 
  2015 & Hungary & 1\% & 0\% & 466 & 0\% & 99\% & 589 \\ 
  2015 & Italy & 2\% & 0\% & 947 & 0\% & 96\% & 1043 \\ 
  2015 & Netherlands & 7\% & 0\% & 521 & 0\% & 83\% & 525 \\ 
  2015 & Poland & 0\% & 0\% &  & 0\% & 97\% & 448 \\ 
  2015 & Spain & 10\% & 0\% & 608 & 0\% & 99\% & 703 \\ 
  2015 & Sweden & 7\% & 2\% & 527 & 0\% & 97\% & 542 \\ 
  2015 & United Kingdom & 2\% & 1\% & 1387 & 0\% & 93\% & 1388 \\ 
  2015 & United States & 10\% & 0\% & 585 & 0\% & 97\% & 544 \\ 
   \hline
\end{tabular}
}

@@

\end{center}
\end{figure}

We start by look at the achievement for some countries. We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled. Some countries have increased their achievement very strongly. For example, France, Austria and surprisingly Sweden. In mathematics, France increase the gap by roughly 0.9. This pattern happens similarly for literacy with an increase of 0.6. For such a short period of time, the magnitud of the increase is reasonably big.

@@

Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider. However, remember that the Y axis is measured in standard deviations. Small changes are actually very huge in practical terms. Take the case of Sweden. The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015. The actual difference is of about 0.6. Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.

The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available. Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.

\end{center}
\end{figure}

Generally speaking, we see that most countries increased their achievement gap. France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing. Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap. In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.

Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase. The previos two graphs for reading are available upon request}. This means that the result is not an artifact of chance. However, it's important to disentangle where is this gap originating from. Is this because the top are improving why the bottom decreases? Or is it that the bottom is catching up? Up next we plot the same graph but show the divergent patterns between high/low SES origins.

{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_increase/decrease-1}

\end{center}
\end{figure}

@@

\end{center}
\end{figure}

@@

In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.

\end{center}
\end{figure}

We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany. We see a negative relationship where the faster the low SES catch up the slower the High SES decrease. Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel). On the other hand, we do see more ocuntries where the both groups are increasing (top right panel). We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.

@@

{\centering \includegraphics[width=8.5in,height=5in]{figure/avgperformance_inequality-1}

@@

\end{center}
\end{figure}

For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}. The correlations are between -0.7343267 and -0.8781625. This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}. However, what they didn't really go in depth was whether this corelation is strongest for some countries over time. With the data available we can look at the evolution separately for many countries.

{\centering \includegraphics[width=5.5in,height=5in]{figure/country_achievement_and_inequality-1}

@@

\end{center}
\end{figure}

In fact, we see a strong correspondence. In countries like France we see quite a functional form between both inequality and performance. As average performance plummets, the achievement gap is increasing. This pattern is also seen in in the Austria, Australia and Germany. The other obvious pattern is the opposite trend. We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present. The surprising case here is the United States. We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well. In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system. However, the explanation behind this pattern is very speculative. Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap. Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
